Efficient pedigree recording for fast population genetics simulation

### 2018-01-10 forwards_paper.tex
Add detailed methods for simupop

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\fwdpy}{\texttt{fwdpy11}}
\newcommand{\cpp}{\texttt{C++}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\ftprime}{\texttt{ftprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
% mbox to avoid hyphenating
\newcommand{\taddnode}[1]{\mbox{\texttt{addnode}}(#1)}
\newcommand{\taddedge}[1]{\mbox{\texttt{addedge}}(#1)}
\newcommand{\tsimplify}[1]{\mbox{\texttt{simplify}}(#1)}
\DeclareMathOperator{\len}{length}
\DeclareMathOperator{\add}{add} % for adding to tables
\newcommand{\lleft}{\mathrm{left}}  % left label
\newcommand{\lright}{\mathrm{right}}
\newcommand{\lid}{\mathrm{id}}

@@

\algstep{W1.}{Initialisation.}{
Set $\Nt \leftarrow \mbox{\texttt{NodeTable}}()$, $\mathcal{E}
\leftarrow \mbox{\texttt{EdgeTable}}()$, $t \leftarrow T$, and $j \leftarrow 0$.
 For $0 \leq k < N$, set $P_k \leftarrow \taddnode{\Nt, T}$.
}

@@

\krt{Are we doing a methods section?}

\subsection{Requirements for the \cpp{} implementation}

\begin{enumerate}
    \item \fwdpy{} 0.13.a0
    \item \fwdpp{} 0.5.7
    \item GNU Scientific Library 1.16
    \item pybind11 2.2.1
    \item GCC 4.8.5
    \item \msprime{} master branch \jda{which commit hash or short version number
		    from pip freeze or {\tt msprime.\_\_version\_\_}?}
\end{enumerate}

All of the above installed via conda, except for msprime, which must be installed from GitHub.

\subsection{Requirements for the \simupop{} implementation}

The simulations are implemented the Python package \ftprime{} 0.0.6rc0
which is available at \url{https://www.github.com/ashander/ftprime}.
Its requirements are encoded in an \texttt{environment.yml} file, compatible
with conda, but \msprime{} must be updated to the version listed below via pip
from GitHub \jda{TODO add command}.
In brief the requirements are:

\begin{enumerate}
    \item \msprime{} \jda{TODO add version from KRT's dev machine}
    \item \simupop{} 1.1.8.3
    \item python 3.5
\end{enumerate}

>>>>

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\fwdpy}{\texttt{fwdpy11}}
\newcommand{\cpp}{\texttt{C++}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\ftprime}{\texttt{ftprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
% mbox to avoid hyphenating
\newcommand{\taddnode}[1]{\mbox{\texttt{addnode}}(#1)}
\newcommand{\taddedge}[1]{\mbox{\texttt{addedge}}(#1)}
\newcommand{\tsimplify}[1]{\mbox{\texttt{simplify}}(#1)}
\newcommand{\tsort}[1]{\mbox{\texttt{sort}}(#1)}
\newcommand{\tnodetable}[1]{\mbox{\texttt{NodeTable}}(#1)}
\newcommand{\tedgetable}[1]{\mbox{\texttt{EdgeTable}}(#1)}
\DeclareMathOperator{\len}{length}
\DeclareMathOperator{\add}{add} % for adding to tables
\newcommand{\lleft}{\mathrm{left}}  % left label
\newcommand{\lright}{\mathrm{right}}
\newcommand{\lid}{\mathrm{id}}

@@

\algstep{W1.}{Initialisation.}{
Set $\Nt \leftarrow \tnodetable()$, $\mathcal{E}
\leftarrow \mbox{\texttt{EdgeTable}}()$, $t \leftarrow T$, and $j \leftarrow 0$.
 For $0 \leq k < N$, set $P_k \leftarrow \taddnode{\Nt, T}$.
}

@@

\subsection*{The \cpp{} implementation}

\paragraph{Requirements.}
The \cpp{} implementation uses several packages installed via conda,
except for \msprime{}, which must be installed from GitHub.

\begin{enumerate}
    \item \fwdpy{} 0.13.a0
    \item \fwdpp{} 0.5.7
    \item GNU Scientific Library 1.16
    \item pybind11 2.2.1
    \item GCC 4.8.5
    \item \msprime{} 0.5.0 \jda{Changed these but the actual sims were not run
		    with this version...}
\end{enumerate}

\subsection*{The \simupop{} implementation}

We implemented simulations in a Python script, usable as a command-line utility,
taking input parameters for population size $N$ and scaled recombination $\rho$.
As described in 'Simulation benchmarks', we set scaled mutation $\theta=\rho$,
we used simplify interval of 1000 and simulated for $T=10N$ generations in each
simulation, introducing deleterious mutations at a rate of $\rho/100$ per
generation.
Because \simupop{} requires explicit locations for loci, we set the number of
loci $n=\rho / (4Nr)$ and marked
$k={\rm round}(\mu / (100 (\mu + \mu / 100))) = 0$ of these as selected loci.
\jda{unfortunately probably what we did...}
The selected loci were chosen as a random sample of all the loci.

We coupled \simupop{} to \msprime{} using a Python class \texttt{RecombCollector} 1)
tracks time, and 2) collects and parses recombination information provided by
the \texttt{Recombinator} class of \simupop{}, adding them to a \tnodetable{} and
\tedgetable{} upon which simplification is performed every 1000 years (using
\texttt{msprime.sort\_tables()} and \texttt{msprime.simplify\_tables}).
The tables are initialized using the \texttt{TreeSequence} returned from
\texttt{msprime.simulate($2N$, Ne=$N$, recombination\_rate=$r/2$, length=$n$)}.
The \texttt{simuPOP.Population()} is initialized and tagged with individual IDs,
a mapping of these IDs to the \tnodetable IDs is passed to
an instance of \texttt{RecombCollector} and used internally to update the tables.
The \texttt{simuPOP.Population()} is then evolved.
Before mating: selected mutations are created (using
\texttt{simuPOP.SNPMutator}) and multiplicative selection occurs (using
\texttt{simuPOP.PyMlSelector}) only on the selected loci;
the internal time of the \texttt{RecombCollector} instance is updated.
During mating: male and female parents are drawn randomly, recombination data is
output by \simupop{} to the \texttt{RecombCollector} instance.
After mating: nothing happens 999/1000 years, every 1000 years the tables are
simplifed.

\texttt{RecombCollector} uses the Python class \texttt{ARGrecorder} internally
to add data to the tables and perform simplification.
This class contains the \tnodetable and \tedgetable and also the mapping from
(haploid) ids to ids in the tables.
It also provides methods (\texttt{add\_individual} and \texttt{add\_record})
to append data to the tables.
The only functions of
\texttt{RecombCollector} not delegated to this internal class are computing
haploid IDs from \simupop{}'s diploid IDs and parsing recomination data into
data suitible for use with use with \texttt{add\_individual} and
\texttt{add\_record}; these functions are used to stream data into the tables as
it is received by an instance of \texttt{RecombCollector}.

\paragraph{Requirements.}
The simulations as implemented depend on the class \texttt{RecombCollector} provided by
the Python package \ftprime{} 0.0.6rc0,
available at \url{https://www.github.com/ashander/ftprime}.
Its requirements are encoded in an \texttt{environment.yml} file, compatible
with conda, but \msprime{} must be updated to the version listed below via pip
from GitHub \jda{TODO add command?}.

In brief the requirements are:

\begin{enumerate}
    \item \msprime{} 0.5.0
    \item \simupop{} 1.1.8.3
    \item python 3.5
\end{enumerate}
